Online supplemental material of the article:\\ \vspace{7mm} 'Hilbert space approximate Bayesian Gaussian processes using Stan'\\ \vspace{7mm} Case Studies

### 2020-03-21 Paper/online_supplemental_material/online_material.tex
updated manuscript_v2, online material and added case studies codes

@@

\title{Online appendices}

@@

\section{Univariate case studies}

\subsection{Same-sex marriage data}\label{ch5_sec_studycaseII}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 observations of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group $i$ ($i=1,\dots,74$). The observational model is a binomial model with parameters population $n_i$ and probability of supporting same-sex marriage $p_i$ per age group $i$,
%
\begin{equation*}
y_i \sim \text{Binomial}(p_i, n_i).
\end{equation*}

\noindent The population per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ or mean number of support people per age group. Probabilities $\bm{p}=(p_1,\dots,p_{74})$ are modeled by a GP function $f:{\rm I\!R} \to {\rm I\!R}$ with a squared exponential covariance function $k$, as a function of age input values $\bm{x}=(x_1,\dots,x_{74})$, and through the {\it logit} function as a link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \text{logit}(f(x_i)) \nonumber \\
f(x) &\sim \mathcal{GP}(0, k(x,x', \theta)).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. In the HSGP model, the function $f(x)$ is approximated as in equation (\ref{eq_approxf}), with the squared exponential spectral density as in equation (\ref{eq_specdens_inf}), and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}).

@@

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 10 times faster than the regular GP, in this particular case.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found in:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-V_Same-sex-marriage
\end{lstlisting}

@@

\section{Multivariate case studies}

\subsection{2D Simulated data}

@@

In the HSGP model with $2$ input dimensions, the latent function $f$, evaluated at input vector $\bm{x} \in {\rm I\!R}^2$, is approximated as in equation (\ref{eq_approxf_multi}), with the $2$-dimensional squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}), and the multivariate eigenfunctions $\phi_j$ and the $2$-vector of eigenvalues $\bm{\lambda}_j$ as in equations (\ref{eq_eigenfunction_multi}) and  (\ref{eq_eigenvalue_multi}), respectively.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the splines model, have been fitted. In all models, the same number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-(left) shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$, and the number of univariate basis functions $m$, for the HSGP model, and knots, for the splines model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-(left), it can be seen a close approximation of the HSGP model to the regular GP model. However, the performance of the splines model is significantly worse. Figure \ref{ch5_fig17_RMSE_exII}-(right) shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computational time.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found in:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-VI_Simulated-data_2D
\end{lstlisting}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig17_time_exII.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 113mm 0mm, clip]{ch5_fig17_legend_exII.pdf}}
\caption{Root mean square error (RMSE) (left) and computational time (right) in seconds per iteration (iteration of the HMC sampling method) of the different methods computed against the data-generating function, as a function of the boundary factor $c$, number of basis functions $m$ and knots.}
  \label{ch5_fig17_RMSE_exII}
\end{figure}

\subsection{Diabetes data}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The matrix $X=[\bm{x}_1\;\bm{x}_2\;\cdots\;\bm{x}_n]^\top \in {\rm I\!R}^{n\times 4}$,  with $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, contains the risk factors, \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \text{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function of the risk factors, which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a Gaussian process with a multivariate squared exponential covariance function $k$ depending on the matrix $X$ of risk factors and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{eqnarray*} \label{ch5_eq_gpprior_gay}
p_i &=& \text{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim& \mathcal{GP}(0, k(\bm{x},\bm{x}, \theta).
\end{eqnarray*}

@@

In the HSGP model with $D$ input dimensions, the function $f(\bm{x})$ evaluated at input vector $\bm{x} \in {\rm I\!R}^D$ is approximated as in equation (\ref{eq_approxf_multi}), with the $D$-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}) and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}), respectively.

@@

In order to assess the performance of the models as a function of the boundary factor, the number of basis functions and knots, different models with different number of basis functions and boundary factor for the HSGP model and different number of knots for the spline model have been fitted. In all models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig19_ELPD_diabetes} shows the expected log predictive density (ELPD; see \cite{vehtari_2012}) as a function of the boundary factor $c$ and the number of univariate basis functions $m$, for the HSGP model, and knots, for the spline model. The ELPD is computed over the actual observations by cross-validation. Basically, with slightly differences, all models show similar performances, due to the fact that the process is very smooth with a relatively very large lengthscale estimate $\ell=4.51$. Even though, a slight pattern of performance improvement can be appreciated as the boundary factor $c$ increases, which fact is because small boundary factors are not allowed when large lengthscales (Figure \ref{fig6_lscale_vs_J_vs_c}).
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig19_ELPD_diabetes.pdf}}
\caption{Expected log predictive density (ELPD) of the different methods as a function of the boundary factor $c$ and the number of basis functions $m$, for the HSGP model, and knots, for the spline model.}
  \label{ch5_fig19_ELPD_diabetes}
\end{figure}

Figure \ref{ch5_fig20_time2D_diabetes} shows the computational times of the different models, regular GP, HSGP and spline, fitted over the dataset, with 2 input dimensions, 3 input dimensions and 4 input dimensions, as a function of the boundary factor $c$ and number of univariate basis functions $m$ and knots. We can appreciate the significant increase of computational time with higher dimensions for the HSGP and spline models. This fact reveals that choosing optimal values for the number of basis functions and boundary factor, looking at the recommendations and diagnosis provided by Figure \ref{fig6_lscale_vs_J_vs_c}, is essential to avoid a excessive computational time, especially in high input dimensionality. It is interesting to be noticed that considering more than 10 knots per dimension in the spline model with 3D is not allowed for an amount of 392 observations. Similarly, just the computation of the input data for the spline model in a 4D input space is computationally very expensive.

The Stan model codes for the exact GP, the approximate GP and the spline models of this case study can be found in:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-III_Diabetes-data
\end{lstlisting}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig20_time2D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 30.2mm 0mm 10mm 0mm, clip]{ch5_fig20_time3D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 30.2mm 0mm 10mm 0mm, clip]{ch5_fig20_time4D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 28mm 50mm 112mm 10mm, clip]{ch5_fig20_legend_diabetes_2.pdf}}
\caption{Time of computation in seconds per iteration (iteration of the HMC sampling method) of the different models fitted over the dataset, with 2 input dimensions (left) and 3 input dimensions (center) and 4 input dimensions (right), as a function of the boundary factor $c$ and number of basis functions $m$, for the HSGP model, and knots, for the spline model.}
  \label{ch5_fig20_time2D_diabetes}
\end{figure}

\subsection{Spatio-temporal land-use classification}\label{ch5_sec_application_caseII}

@@

\noindent which is related to the vector of probabilities $\bm{p}_{it}$ through the 'softmax' link function (equation \ref{ch2_eq_softmax} in chapter \ref{ch2_bayes}):
%
\begin{eqnarray*}
\bm{p}_{it} = \text{softmax}(f(\bm{x}_{it})).
\end{eqnarray*}

Each individual function $f_k(\bm{x})$ is modeled as a first-order additive model plus the second-order additive effects between time input variable ($x^D$) and all the other inputs as follows: 
%
\begin{eqnarray} \label{ch5_eq_f_parcels}
f_k(\bm{x}) = \sum_{d=1}^{D} g_d(x^d) + \sum_{d=1}^{D-1} h_{d,D}(x^d,x^D).
%\mathcal{HSGP}(x^d, S(\bm{\lambda}), \theta_d),
\end{eqnarray}

\noindent The first-order components $\{g_d(x^d)\}_{d=1}^D$ in equation (\ref{ch5_eq_f_parcels}) are modeled as unidimensional HSGP models:
%
\begin{eqnarray*}
g_d(x^d) \sim \mathcal{HSGP}(x^d, S, \theta_{d,1}).
\end{eqnarray*}

\noindent In the HSGP model, a first-order components $g_d(x^d)$, evaluated at input value $x^d \in {\rm I\!R}$, is approximated as in equation (\ref{eq_approxf}) with the squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}) and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively.

The second-order components $\{h_{d,D}(x^d,x^D)\}_{d=1}^{D-1}$ in equation (\ref{ch5_eq_f_parcels}) are modeled as two-dimensional HSGP models:
%
\begin{eqnarray*}
h_{d,D}(x^d,x^D) \sim \mathcal{HSGP}(x^d,x^D, S, \theta_{d,D}).
\end{eqnarray*}

\noindent In the HSGP model, a second-order component $h_{d,D}(x^d,x^D)$, evaluated at inputs $x^d \in {\rm I\!R}$ and $x^D \in {\rm I\!R}$, is approximated as in equation (\ref{eq_approxf_multi}) with the two-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}) and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}), respectively.

@@

In the case of the first-order components, the normalized lengthscale estimates $\left(\frac{2\cdot \hat{\ell}_{d,1}}{|x^d_{max}-x^d_{min}|}\right)$ are all bigger than the minimum lengthscale reported by Figure \ref{fig6_lscale_vs_J_vs_c} as a function of $m$, $c$. Which means that the used number of basis functions ($m=15$) and boundary factor ($c=2.5$) are suitable values for modeling accurately the input effects.

@@

The Stan model code for the approximate GP model of this case study can be found in:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-IV_Land-use-classification
\end{lstlisting}

>>>>

\usepackage{chngcntr}

@@

\title{Online supplemental material of the article:\\ \vspace{7mm} 'Hilbert space approximate Bayesian Gaussian processes using Stan'\\ \vspace{7mm}
Case Studies}

\author{TODO}

\date{}

@@

\counterwithin{figure}{section}
\counterwithin{table}{section}
\counterwithin{equation}{section}

\section{Same-sex marriage case study}\label{ch5_sec_studycaseII}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 observations of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group $i$ ($i=1,\dots,74$). The observational model is a binomial model with parameters population $n_i$ and probability of supporting same-sex marriage $p_i$ per age group $i$,
%
\begin{equation*}
y_i \sim \text{Binomial}(p_i, n_i).
\end{equation*}

\noindent The population per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ or mean number of support people per age group. Probabilities $\bm{p}=(p_1,\dots,p_{74})$ are modeled by a GP function $f:{\rm I\!R} \to {\rm I\!R}$ with a squared exponential covariance function $k$, as a function of age input values $\bm{x}=(x_1,\dots,x_{74})$, and through the {\it logit} function as a link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(x_i)) \nonumber \\
f(x) &\sim \mathcal{GP}(0, k(x,x', \theta)).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. In the HSGP model, the function $f(x)$ is approximated as in equation (7) %(\ref{eq_approxf})
, with the squared exponential spectral density as in equation (1)%(\ref{eq_specdens_inf})
, and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (5) %(\ref{eq_eigenvalue}) 
and (6)%(\ref{eq_eigenfunction})
.

@@

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 15 times faster than the regular GP and 5 times faster than the spline model, for this particular case and univariate input space. {\color{blue}The increase of computation as function of the number of basis functions in a univariate input space is relatively slight, as can be seen in the figure}.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data}.

%\begin{lstlisting}[breaklines]
%https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data
%\end{lstlisting}

@@

\section{Case study of simulated data for a bivariate function}

@@

In the HSGP model with $2$ input dimensions, the latent function $f$, evaluated at input vector $\bm{x} \in {\rm I\!R}^2$, is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $2$-dimensional squared exponential spectral density $S$ as in equation (1)% (\ref{eq_specdens_inf})
, and the multivariate eigenfunctions $\phi_j$ and the $2$-vector of eigenvalues $\bm{\lambda}_j$ as in equations (10) %(\ref{eq_eigenfunction_multi})
 and (9)% (\ref{eq_eigenvalue_multi})
 , respectively.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the splines model, have been fitted. In all models, the same number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$, and the number of univariate basis functions $m$, for the HSGP model, and knots, for the splines model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen a close approximation of the HSGP model to the regular GP model. However, the performance of the splines model is significantly worse. Figure \ref{ch5_fig17_RMSE_exII}-right shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computational time. {\color{blue}Even though in a bivariate input space the computation demand increases significantly with the number of dimensions or knots, either the HSGP or spline models work significantly better than regular GP, even for highly wiggly functions where a high number of basis functions or knots are required in the approximations. However, severe difficulties have been found in building the spline model with $50$ knots for this bivariate function}.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig17_time_exII.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 112mm 0mm, clip]{ch5_fig17_legend_exII.pdf}}
\caption{Root mean square error (RMSE) (left) and computational time (right) in seconds per iteration (iteration of the HMC sampling method) of the different methods computed against the data-generating function, as a function of the boundary factor $c$, number of basis functions $m$ and knots.}
  \label{ch5_fig17_RMSE_exII}
\end{figure}

\section{Diabetes case study}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The matrix $X=[\bm{x}_1\;\bm{x}_2\;\cdots\;\bm{x}_n]^\top \in {\rm I\!R}^{n\times 4}$,  with $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, contains the risk factors, \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \text{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function of the risk factors, which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a Gaussian process with a multivariate squared exponential covariance function $k$ depending on the matrix $X$ of risk factors and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x},\bm{x}, \theta).
\end{align*}

@@

In the HSGP model with $D$ input dimensions, the function $f(\bm{x})$ evaluated at input vector $\bm{x} \in {\rm I\!R}^D$ is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $D$-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (1) %(\ref{eq_specdens_inf}) 
and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (9) % (\ref{eq_eigenvalue_multi}) 
and (10)% (\ref{eq_eigenfunction_multi})
, respectively.

@@

In order to assess the performance of the models as a function of the boundary factor, the number of basis functions and knots, different models with different number of basis functions and boundary factor for the HSGP model and different number of knots for the spline model have been fitted. In all models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig19_ELPD_diabetes} shows the expected log predictive density (ELPD; see \cite{vehtari_2012}) as a function of the boundary factor $c$ and the number of univariate basis functions $m$, for the HSGP model, and knots, for the spline model. The ELPD is computed over the actual observations by cross-validation. Basically, with slightly differences, all models show similar performances, due to the fact that the process is very smooth with a relatively very large lengthscale estimate $\ell=4.51$. Even though, a slight pattern of performance improvement can be appreciated as the boundary factor $c$ increases, which fact is because small boundary factors are not allowed when large lengthscales (Figure 6% \ref{fig6_lscale_vs_J_vs_c}
).
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig19_ELPD_diabetes.pdf}}
\caption{Expected log predictive density (ELPD) of the different methods as a function of the boundary factor $c$ and the number of basis functions $m$, for the HSGP model, and knots, for the spline model.}
  \label{ch5_fig19_ELPD_diabetes}
\end{figure}

Figure \ref{ch5_fig20_time2D_diabetes} shows the computational times of the different models, regular GP, HSGP and spline, fitted over the dataset, with 2 input dimensions, 3 input dimensions and 4 input dimensions, as a function of the boundary factor $c$ and number of univariate basis functions $m$ and knots. We can appreciate the significant increase of computational time with higher dimensions for the HSGP and spline models. In this sense, choosing optimal values for the number of basis functions and boundary factor, looking at the recommendations and diagnosis provided by Figure 6% \ref{fig6_lscale_vs_J_vs_c}
, can be essential to avoid a excessive computational time, especially in multivariate input spaces. It is interesting to be noticed that considering more than 10 knots per dimension in the spline model with 3D is not allowed for an amount of 392 observations. Similarly, just the computation of the input data for the spline model in a 4D input space is computationally very expensive.

The Stan model codes for the exact GP, the approximate GP and the spline models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Diabetes-data}.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig20_time2D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 30.2mm 0mm 10mm 0mm, clip]{ch5_fig20_time3D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 30.2mm 0mm 10mm 0mm, clip]{ch5_fig20_time4D_diabetes_2.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 28mm 50mm 112mm 10mm, clip]{ch5_fig20_legend_diabetes_2.pdf}}
\caption{Time of computation in seconds per iteration (iteration of the HMC sampling method) of the different models fitted over the dataset, with 2 input dimensions (left) and 3 input dimensions (center) and 4 input dimensions (right), as a function of the boundary factor $c$ and number of basis functions $m$, for the HSGP model, and knots, for the spline model. The y-axis is on a logarithmic scale.}
  \label{ch5_fig20_time2D_diabetes}
\end{figure}

\section{Spatio-temporal land-use classification case study}\label{ch5_sec_application_caseII}

@@

\noindent which is related to the vector of probabilities $\bm{p}_{it}$ through the 'softmax' link function:
%
\begin{equation*}
\bm{p}_{it} = \mathrm{softmax}(f(\bm{x}_{it})).
\end{equation*}

Each individual function $f_k(\bm{x})$ is modeled as a first-order additive model plus the second-order additive effects between time input variable ($x^D$) and all the other inputs as follows: 
%
\begin{equation} \label{ch5_eq_f_parcels}
f_k(\bm{x}) = \sum_{d=1}^{D} g_d(x^d) + \sum_{d=1}^{D-1} h_{d,D}(x^d,x^D).
%\mathcal{HSGP}(x^d, S(\bm{\lambda}), \theta_d),
\end{equation}

\noindent The first-order components $\{g_d(x^d)\}_{d=1}^D$ in equation (\ref{ch5_eq_f_parcels}) are modeled as unidimensional HSGP models:
%
\begin{equation*}
g_d(x^d) \sim \mathcal{HSGP}(x^d, S, \theta_{d,1}).
\end{equation*}

\noindent In the HSGP model, a first-order components $g_d(x^d)$, evaluated at input value $x^d \in {\rm I\!R}$, is approximated as in equation (7) % (\ref{eq_approxf}) 
with the squared exponential spectral density $S$ as in equation (1) % (\ref{eq_specdens_inf})
 and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (5) % (\ref{eq_eigenvalue})
and (6)% (\ref{eq_eigenfunction})
  , respectively.

The second-order components $\{h_{d,D}(x^d,x^D)\}_{d=1}^{D-1}$ in equation (\ref{ch5_eq_f_parcels}) are modeled as two-dimensional HSGP models:
%
\begin{equation*}
h_{d,D}(x^d,x^D) \sim \mathcal{HSGP}(x^d,x^D, S, \theta_{d,D}).
\end{equation*}

\noindent In the HSGP model, a second-order component $h_{d,D}(x^d,x^D)$, evaluated at inputs $x^d \in {\rm I\!R}$ and $x^D \in {\rm I\!R}$, is approximated as in equation (12)% (\ref{eq_approxf_multi})
 with the two-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (1) % (\ref{eq_specdens_inf}) 
 and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (9) % (\ref{eq_eigenvalue_multi}) 
 and (10)% (\ref{eq_eigenfunction_multi})
 , respectively.

@@

In the case of the first-order components, the normalized lengthscale estimates $\left(\frac{2\cdot \hat{\ell}_{d,1}}{|x^d_{max}-x^d_{min}|}\right)$ are all bigger than the minimum lengthscale reported by Figure (6) % \ref{fig6_lscale_vs_J_vs_c} 
as a function of $m$, $c$. Which means that the used number of basis functions ($m=15$) and boundary factor ($c=2.5$) are suitable values for modeling accurately the input effects.

@@

The Stan model code for the approximate GP model of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Land-use-classification}.
